Inferring true changes in microbial abundance from taxonomically-biased microbiome measurements

### 2021-11-10 real-world-implications.Rmd
Add and rearrange figures
- Add 1 main and 1 supplemental MOMSPI figure
- Move supplemental MOMSPI bias-sensitivity-demo figures to a new
  supplemental figures appendix

One approach to studying the mean efficiency empirically is to analyze data from studies for which control measurements allow direct measurement of species' efficiencies.
To characterize bias in a 16S sequencing protocol developed for the Vaginal Human Microbiome Project (VaHMP), @brooks2015thet performed 16S rRNA gene sequencing of mock communities of seven bacterial species thought to play a critical role in the human vaginal microbiome.
This protocol was later used for the VaHMP's Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study (@fettweis2019thev).
We used the method for measuring and correcting bias developed by @mclaren2019cons to measure bias from the mock communities and analyze its potential impact on DA analyses that use the MOMSPI measurements (SI Analysis MOMS-PI).
The vaginal microbiome during pregnancy is characterized by relatively low diversity, with a single species often forming a vast majority of calibrated or uncalibrated read counts.
The three species that most often dominate the calibrated profiles are _Gardnerella vaginalis_, _Lactobacillus iners_, and _Lactobacillus crispatus_.
Our mock-community analysis indicate _G. vaginalis_ has the lowest efficiency of the seven species, roughly 20X lower than _L. crispatus_ and 30X lower than _L. iners_ (the most efficiently measured).
As a result, the mean efficiency varies substantially across samples, with samples dominated by a _Lactobacillus_ species typically having an efficiency that is 3-20X greater than samples dominated by _Gardnerella_ (SI figure).
Shifts between _Lactobacillus_ and _Gardnerella_ dominance are common in between-women comparisons, and also occasionally occur between subsequent sampling points in individual women (SI Figure).
Examination of the multiplicative trajectories in the proportions of individual species indicates that such transitions can cause spurious fold changes in lower-abundance species (SI Figure).
Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.

>>>>

One approach to studying the mean efficiency empirically is to analyze data from studies for which control measurements allow direct measurement of species' efficiencies.
To characterize bias in a 16S sequencing protocol developed for the Vaginal Human Microbiome Project (VaHMP), @brooks2015thet performed 16S rRNA gene sequencing of mock communities of seven bacterial species thought to play a critical role in the human vaginal microbiome.
This protocol was later used for the VaHMP's Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study (@fettweis2019thev).
We used the method for measuring and correcting bias developed by @mclaren2019cons to measure bias from the mock communities and analyze its potential impact on DA analyses that use the MOMSPI measurements (SI Analysis MOMS-PI).
The vaginal microbiome during pregnancy is characterized by relatively low diversity, with a single species often forming a vast majority of calibrated or uncalibrated read counts.
The three species that most often dominate the calibrated profiles are _Gardnerella vaginalis_, _Lactobacillus iners_, and _Lactobacillus crispatus_.
Our mock-community analysis indicate _G. vaginalis_ has the lowest efficiency of the seven species, roughly 20X lower than _L. crispatus_ and 30X lower than _L. iners_ (the most efficiently measured).
As a result, the mean efficiency varies substantially across samples, with samples dominated by a _Lactobacillus_ species typically having an efficiency that is 3-20X greater than samples dominated by _Gardnerella_ (Figure \@ref(fig:momspi-mean-efficiency-dist)).
Shifts between _Lactobacillus_ and _Gardnerella_ dominance are common in between-women comparisons, and also occasionally occur between subsequent sampling points in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)).
Examination of the multiplicative trajectories in the proportions of individual species indicates that such transitions can cause spurious fold changes in lower-abundance species (SI Figure TODO).
Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.

<!-- begin figure -->

```{r momspi-mean-efficiency-dist, fig.cap = '(ref:cap-momspi-mean-efficiency-dist)'}
svg_path <- fs::path(
  "notebook/_posts/2021-11-01-momspi-summary/momspi-summary_files", 
  "figure-html5/momspi-mean-efficiency-dist-1.svg"
)
if (knitr::is_html_output()) {
  fig_path <- svg_path
} else if (knitr::is_latex_output()) {
  fig_path <- svg_to_pdf(svg_path)
}
knitr::include_graphics(fig_path)
rm(svg_path, fig_path)
```

(ref:cap-momspi-mean-efficiency-dist) **The mean efficiency in vaginal samples from the MOMS-PI study varies with the most abundant species.**

<!-- end figure -->

### 2021-11-10 solutions.Rmd
Add and rearrange figures
- Add 1 main and 1 supplemental MOMSPI figure
- Move supplemental MOMSPI bias-sensitivity-demo figures to a new
  supplemental figures appendix

A bias-sensitivity analysis examines how sensitive the results of a given DA analysis are to the working assumption about the bias of a given protocol.
This working assumption is typically is that there is no bias, but could also be bias measured in a previous experiment or predicted from taxonomic features like 16S copy number.
A straightforward approach to bias-sensitivity analysis is to use computer simulation to re-analyze the MGS data across a set of simulated taxonomic biases.
First, multiple sets of efficiency vectors are randomly generated, each representing a possible quantification of taxonomic bias for the study.
Second, each efficiency vector is used to calibrate the original MGS measurements; each set of calibrated measurements represents the true community compositions under the given hypothesized bias. 
Finally, the DA analysis is re-run on each set of calibrated measurements and the distribution of results is compared graphically or with summary statistics.
A graphical summary of such an analysis being applied to results from the corncob frequentist DA- analysis tool 'corncob' are shown in \@ref(fig:sensitivity-example).
The simulated efficiency vectors may be generated to different hypotheses about bias in the given system, such as its overall magnitude and correlation among phylogenetically-related species.
Thus the utility of such simulations will increase the more we learn about the properties of taxonomic bias in different systems.

@@

## Supplemental figures

```{r sensitivity-example, fig.cap = '(ref:cap-sensitivity-example)'}
svg_path <- here::here(
  "notebook/_posts/2021-10-18-evaluate-robustness-example/",
  "evaluate-robustness-example_files/figure-html5/summary_plot-1.svg"
)
if (knitr::is_html_output()) {
  fig_path <- svg_path
} else if (knitr::is_latex_output()) {
  fig_path <- svg_to_pdf(svg_path)
}
knitr::include_graphics(fig_path)
rm(svg_path, fig_path)
```

(ref:cap-sensitivity-example) **A bias-sensivity analysis can be performed to examine how sensitive the results of a DA analysis are to assumptions about taxonomic bias in community measurements.** The figure shows the results of a bias-sensitivity analysis used to study the effect of bias on the association of _Gardnerella vaginalis_ and preterm birth that was investigated by @callahan2017repl. 100 random efficiency vectors were drawn at 6 different bias strengths (quantified by the variance in log efficiency, $\sigma_{e}^{2}$). Each efficiency vector was used to calibrate the MGS profiles and perform a DA association test of _G. vaginalis_ versus the host's preterm birth outcome; regression coefficients $\hat \beta$ indicate the increase of average logit proportion of  _G. vaginalis_ in women who experienced preterm birth.

>>>>

A bias-sensitivity analysis examines how sensitive the results of a given DA analysis are to the working assumption about the bias of a given protocol.
This working assumption is typically is that there is no bias, but could also be bias measured in a previous experiment or predicted from taxonomic features like 16S copy number.
A straightforward approach to bias-sensitivity analysis is to use computer simulation to re-analyze the MGS data across a set of simulated taxonomic biases.
First, multiple sets of efficiency vectors are randomly generated, each representing a possible quantification of taxonomic bias for the study.
Second, each efficiency vector is used to calibrate the original MGS measurements; each set of calibrated measurements represents the true community compositions under the given hypothesized bias. 
Finally, the DA analysis is re-run on each set of calibrated measurements and the distribution of results is compared graphically or with summary statistics.
A graphical summary of such an analysis being applied to results from the corncob frequentist DA- analysis tool 'corncob' are shown in SI Figure \@ref(fig:sensitivity-example).
The simulated efficiency vectors may be generated to different hypotheses about bias in the given system, such as its overall magnitude and correlation among phylogenetically-related species.
Thus the utility of such simulations will increase the more we learn about the properties of taxonomic bias in different systems.

@@
